Faster generic IND-CCA2 secure KEM using ``encrypt-then-MAC''

### 2024-10-04 etm-transform/manuscript/main.tex
Minor fixes for the body

Where the expectation is taken with respect to the probability distribution of all possible keypairs $(\pk, \sk) \leftsample \texttt{PKE.KeyGen()}$. For many lattice-based cryptosystems, including ML-KEM, decryption failures could leak information about the secret key, although the probability of a decryption failure is low enough that classical adversaries cannot exploit decryption failure more than they can defeat the underlying lattice problem (see table \ref{tbl:ml-kem-dec-failure-probs}).

\begin{table}[H]
    \centering

    \begin{tabular}{|c|c|c|c|}
        \hline
        & ML-KEM-512 & ML-KEM-768 & ML-KEM-1024 \\
        \hline
        $\delta$ & ??? & ??? & ??? \\
        \hline
    \end{tabular}
    \caption{Estimated probability of decryption failure in ML-KEM}\label{tbl:ml-kem-dec-failure-probs}
\end{table}

On the other hand, a quantum adversary may be able to exploit decryption failure in reasonable runtime by efficiently searching through all possible inputs using Grover's search algorithm. For that, ML-KEM made slight modifications in its KEM construction to prevent quantum adversary from precomputing large lookup table. We refer readers to \cite{avanzi2019crystals} and \cite{bos2018crystals} for the details.

@@

\begin{figure}[H]
    \centering
    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\keygen()$}
            \begin{algorithmic}[1]
                \State $(\pk, \sk^\prime) \leftsample \pke\texttt{.}\keygen()$
                \State $z \leftsample \mathcal{M}$
                \State $\sk \leftarrow \sk^\prime \Vert z$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\encap(\pk)$}
            \begin{algorithmic}[1]
                \Ensure $\pk$ is some $\pke$ public key
                \State $m \leftsample \mathcal{M}$
                \State $k \leftarrow G(m)$
                \State $c^\prime \leftsample \pke\texttt{.}\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow H(m, c^\prime)$
                \State $c \leftarrow c^\prime \Vert t$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\decap(\sk, c)$}
            \begin{algorithmic}[1]
                \Require $c = c^\prime \Vert t, \sk = sk^\prime \Vert z$
                \Ensure $c^\prime$ is some $\pke$ ciphertext
                \Ensure $t$ is some $\mac$ tag
                \Ensure $\sk^\prime$ is some $\pke$ secret key
                \Ensure $z$ is some $\pke$ plaintext
                \State $(c^\prime, t) \leftarrow c$
                \State $(\sk^\prime, z) \leftarrow \sk$
                \State $\hat{m} \leftarrow \pke\texttt{.}\decrypt(\sk^\prime, c^\prime)$
                \State $\hat{k} \leftarrow G(\hat{m})$
                \If{$\mac(\hat{k}, c^\prime) \neq t$}
                    \State $K \leftarrow H(z, c^\prime)$
                \Else
                    \State $K \leftarrow H(\hat{m}, c^\prime)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{$\kem_\etm$ routines}\label{fig:etm-routines} 
\end{figure}

@@

A detailed description of $\texttt{K-PKE}$'s routines can be found in FIPS 203 (TODO: citation). The ``encrypt-then-MAC'' routines are described in figure \ref{fig:ml-kem-plus-routines}.

@@

CMAC is based on the CBC-MAC with the block cipher instantiated from AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time. A summary of the computation can be found in figure \ref{fig:aes-256-cmac}

\begin{figure}[h]
    \centering
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{Sub-key derivation}
            \begin{algorithmic}[1]
                \Require 256-bit key $k$
                \Require $\texttt{const\_Rb = 0x87}$
                \State $l \leftarrow \texttt{AES-256}(k, 0^{128})$
                \If{$\texttt{MostSignificantBit}(l) = 0$}
                    \State $k_1 \leftarrow \texttt{l {<}{<} 1}$
                \Else 
                    \State $k_1 \leftarrow \texttt{l {<}{<} 1} \oplus \texttt{const\_Rb}$
                \EndIf
                \If{$\texttt{MostSignificantBit}(k_1) = 0$}
                    \State $k_2 \leftarrow k_1 \texttt{ {<}{<} 1}$
                \Else 
                    \State $k_2 \leftarrow k_1 \texttt{ {<}{<} 1} \oplus \texttt{const\_Rb}$
                \EndIf
                \State \Return $k_1, k_2$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{CMAC(k, m)}
            \begin{algorithmic}[1]
                \Require 256-bit symmetric key $k$
                \State $(k_1, k_2) \leftarrow \texttt{deriveSubKey}(k)$
                \State $n \leftarrow \left\lceil \texttt{bytesLen}(m) / 16\right\rceil$
                \If{$n = 0$}
                    \State $n \leftarrow 1$
                    \State $m_\texttt{last} \leftarrow m_n \oplus k_2$
                \ElsIf{$\texttt{bytesLen(m)} \mod 16 = 0$}
                    \State $m_\texttt{last} \leftarrow m_n \oplus k_1$
                \Else 
                    \State $m_\texttt{last} \leftarrow m_n \oplus k_2$
                \EndIf

                \State $x = 0^{128}$
                \For{$i \in \{1, 2, \ldots, n-1\}$}
                    \State $y \leftarrow x \oplus m_i$
                    \State $x \leftarrow \texttt{AES-256}(k, y)$
                \EndFor
                \State $y \leftarrow m_\texttt{last} \oplus x$
                \State $t \leftarrow \texttt{AES-256}(k, y)$
                \State \Return $t$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{AES-256 CMAC}\label{fig:aes-256-cmac}
\end{figure}

@@

@@

Since \texttt{K-PKE.\encrypt} carries significantly more computational complexity than \texttt{K-PKE.\decrypt} or any MAC we chose, the performance advantage of the ``encrypt-then-MAC'' transformation over the $U^{\not\bot}_m$ transformation is dominated by the runtime saving gained from replacing \emph{re-encryption} with MAC. A comparison between ML-KEM and variations of the ML-KEM-ETM can be found in table \ref{tbl:kem-performance}

@@

    \begin{tabular}[t]{|cc|}
        \hline
        % TODO: this title needs to show that parameters are identical across all variants
        \multicolumn{2}{|c|}{\bf 128-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 800 \\
        $\sk$ size & 1632 \\
        \texttt{ct} size & 768 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 75945 \\
        Average & 76171 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \texttt{ML-KEM-512} & 91467 & 92065 & 121185 & 121650 \\
        \hline
        \texttt{Kyber512} & 97811 & 98090 & 119937 & 120299 \\
        \hline
        % ML-KEM+-512
        $\texttt{ML-KEM-512}^+$ w/ Poly1305 & 93157 & 93626 & 33733 & 33908 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ GMAC & 97369 & 97766 & 37725 & 37831 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ CMAC & 99739 & 99959 & 40117 & 39943 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ KMAC & 101009 & 101313 & 40741 & 40916 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 192-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1184 \\
        $\sk$ size & 2400 \\
        \texttt{ct} size & 1088 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 129895 \\
        Average & 130650 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \texttt{ML-KEM-768} & 136405 & 147400 & 186445 & 187529 \\
        \hline
        \texttt{Kyber768} & 153061 & 153670 & 182129 & 182755 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ Poly1305 & 146405 & 146860 & 43315 & 43463 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ GMAC & 149525 & 150128 & 46513 & 46706 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ CMAC & 153139 & 153735 & 49841 & 50074 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ KMAC & 155219 & 155848 & 52415 & 52611 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 256-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1568 \\
        $\sk$ size & 3168 \\
        \texttt{ct} size & 1568 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 194921 \\
        Average & 195465 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \texttt{ML-KEM-1024} & 199185 & 199903 & 246245 & 247320 \\
        \hline
        \texttt{Kyber1024} & 222351 & 223260 & 258231 & 259067 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ Poly1305 & 205763 & 206499 & 51375 & 51562 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ GMAC & 208805 & 209681 & 54573 & 54780 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ CMAC & 213667 & 214483 & 59175 & 59408 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ KMAC & 216761 & 217468 & 62269 & 62516 \\
        \hline
    \end{tabular}
\end{table}

>>>>

Where the expectation is taken with respect to the probability distribution of all possible keypairs $(\pk, \sk) \leftsample \texttt{PKE.KeyGen()}$. For many lattice-based cryptosystems, including ML-KEM, decryption failures could leak information about the secret key, although the probability of a decryption failure is low enough that classical adversaries cannot exploit decryption failure more than they can defeat the underlying lattice problem. On the other hand, a quantum adversary may be able to exploit decryption failure in reasonable runtime by efficiently searching through all possible inputs using Grover's search algorithm. For that, ML-KEM made slight modifications in its KEM construction to prevent quantum adversary from precomputing large lookup table. We refer readers to \cite{avanzi2019crystals} and \cite{bos2018crystals} for the details.

@@

\begin{figure}[H]
    \centering
    \begin{minipage}[t]{0.5\textwidth}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\keygen()$}
            \begin{algorithmic}[1]
                \State $(\pk, \sk^\prime) \leftsample \pke\texttt{.}\keygen()$
                \State $z \leftsample \mathcal{M}$
                \State $\sk \leftarrow \sk^\prime \Vert z$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\encap(\pk)$}
            \begin{algorithmic}[1]
                \State $m \leftsample \mathcal{M}$
                \State $k \leftarrow G(m)$
                \State $c^\prime \leftsample \pke\texttt{.}\encrypt(\pk, m)$
                \State $t \leftarrow \mac(k, c^\prime)$
                \State $K \leftarrow H(m, c^\prime)$
                \State $c \leftarrow c^\prime \Vert t$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}\hfill
    \begin{minipage}[t]{0.49\textwidth}
        \begin{algorithm}[H]
            \caption*{$\kem_\etm.\decap(\sk, c)$}
            \begin{algorithmic}[1]
                \State $(c^\prime, t) \leftarrow c$
                \State $(\sk^\prime, z) \leftarrow \sk$
                \State $\hat{m} \leftarrow \pke\texttt{.}\decrypt(\sk^\prime, c^\prime)$
                \State $\hat{k} \leftarrow G(\hat{m})$
                \If{$\mac(\hat{k}, c^\prime) \neq t$}
                    \State $K \leftarrow H(z, c^\prime)$
                \Else
                    \State $K \leftarrow H(\hat{m}, c^\prime)$
                \EndIf
                \State \Return $K$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{$\kem_\etm$ routines}\label{fig:etm-routines} 
\end{figure}

@@

We refer readers to \cite{FIPS203} for the details of the \texttt{K-PKE} routines. The ``encrypt-then-MAC'' KEM routines are described in figure \ref{fig:ml-kem-plus-routines} below.

@@

CMAC is based on the CBC-MAC with the block cipher instantiated from AES-256. To compute a CMAC tag, the message is first broke into 128-bit blocks with appropriate padding. Each block is first XOR'd with the previous block's output, then encrypted under AES using the symmetric key. The final output is XOR'd with a sub key derived from the symmetric key, before being encrypted for one last time. %A summary of the computation can be found in figure \ref{fig:aes-256-cmac}

% \begin{figure}[h]
%     \centering
%     \begin{minipage}[t]{0.49\textwidth}
%         \begin{algorithm}[H]
%             \caption*{Sub-key derivation}
%             \begin{algorithmic}[1]
%                 \Require 256-bit key $k$
%                 \Require $\texttt{const\_Rb = 0x87}$
%                 \State $l \leftarrow \texttt{AES-256}(k, 0^{128})$
%                 \If{$\texttt{MostSignificantBit}(l) = 0$}
%                     \State $k_1 \leftarrow \texttt{l {<}{<} 1}$
%                 \Else 
%                     \State $k_1 \leftarrow \texttt{l {<}{<} 1} \oplus \texttt{const\_Rb}$
%                 \EndIf
%                 \If{$\texttt{MostSignificantBit}(k_1) = 0$}
%                     \State $k_2 \leftarrow k_1 \texttt{ {<}{<} 1}$
%                 \Else 
%                     \State $k_2 \leftarrow k_1 \texttt{ {<}{<} 1} \oplus \texttt{const\_Rb}$
%                 \EndIf
%                 \State \Return $k_1, k_2$
%             \end{algorithmic}
%         \end{algorithm}
%     \end{minipage}\hfill
%     \begin{minipage}[t]{0.49\textwidth}
%         \begin{algorithm}[H]
%             \caption*{CMAC(k, m)}
%             \begin{algorithmic}[1]
%                 \Require 256-bit symmetric key $k$
%                 \State $(k_1, k_2) \leftarrow \texttt{deriveSubKey}(k)$
%                 \State $n \leftarrow \left\lceil \texttt{bytesLen}(m) / 16\right\rceil$
%                 \If{$n = 0$}
%                     \State $n \leftarrow 1$
%                     \State $m_\texttt{last} \leftarrow m_n \oplus k_2$
%                 \ElsIf{$\texttt{bytesLen(m)} \mod 16 = 0$}
%                     \State $m_\texttt{last} \leftarrow m_n \oplus k_1$
%                 \Else 
%                     \State $m_\texttt{last} \leftarrow m_n \oplus k_2$
%                 \EndIf

%                 \State $x = 0^{128}$
%                 \For{$i \in \{1, 2, \ldots, n-1\}$}
%                     \State $y \leftarrow x \oplus m_i$
%                     \State $x \leftarrow \texttt{AES-256}(k, y)$
%                 \EndFor
%                 \State $y \leftarrow m_\texttt{last} \oplus x$
%                 \State $t \leftarrow \texttt{AES-256}(k, y)$
%                 \State \Return $t$
%             \end{algorithmic}
%         \end{algorithm}
%     \end{minipage}
%     \caption{AES-256 CMAC}\label{fig:aes-256-cmac}
% \end{figure}

@@

From our testing, we found Poly1305 to exhibit the best performance characteristics. However, there are additional security considerations that may require the use of other less efficient MAC instances. For example, it is possible for an adversary with large computing infrastructure or quantum computers to pre-compute a large lookup table mapping symmetric key to the source plaintext. Upon receiving a ciphertext, the adversary can then search through the lookup table for a matching key, which would've revealed the corresponding decryption. We partially mitigated such attack by deriving the symmetric key from both the public key and the plaintext, but in case of a long-term keypair, the adversary might still be able to compute a large lookup table AFTER obtaining the long-term public key. Further mitigation could include using larger-size keys, which can be accomplished either by using a Carter-Wegman MAC that operates on a larger finite field or using a MAC with a variable key-length such as KMAC.

@@

Since \texttt{K-PKE.\encrypt} carries significantly more computational complexity than \texttt{K-PKE.\decrypt} or any MAC we chose, the performance advantage of the ``encrypt-then-MAC'' transformation over the $U^{\not\bot}_m$ transformation is dominated by the runtime saving gained from replacing \emph{re-encryption} with MAC. A comparison between ML-KEM and variations of the $\text{ML-KEM}^+$ can be found in table \ref{tbl:kem-performance}

@@

    \begin{tabular}[t]{|cc|}
        \hline
        % TODO: this title needs to show that parameters are identical across all variants
        \multicolumn{2}{|c|}{\bf 128-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 800 \\
        $\sk$ size & 1632 \\
        \text{ct} size & 768 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 75945 \\
        Average & 76171 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-512} & 91467 & 92065 & 121185 & 121650 \\
        \hline
        \text{Kyber512} & 97811 & 98090 & 119937 & 120299 \\
        \hline
        % ML-KEM+-512
        $\text{ML-KEM}^+$-512 w/ Poly1305 & 93157 & 93626 & 33733 & 33908 \\
        \hline
        $\text{ML-KEM}^+$-512 w/ GMAC & 97369 & 97766 & 37725 & 37831 \\
        \hline
        $\text{ML-KEM}^+$-512 w/ CMAC & 99739 & 99959 & 40117 & 39943 \\
        \hline
        $\text{ML-KEM}^+$-512 w/ KMAC & 101009 & 101313 & 40741 & 40916 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 192-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1184 \\
        $\sk$ size & 2400 \\
        \text{ct} size & 1088 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 129895 \\
        Average & 130650 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-768} & 136405 & 147400 & 186445 & 187529 \\
        \hline
        \text{Kyber768} & 153061 & 153670 & 182129 & 182755 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ Poly1305 & 146405 & 146860 & 43315 & 43463 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ GMAC & 149525 & 150128 & 46513 & 46706 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ CMAC & 153139 & 153735 & 49841 & 50074 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ KMAC & 155219 & 155848 & 52415 & 52611 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 256-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1568 \\
        $\sk$ size & 3168 \\
        \text{ct} size & 1568 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 194921 \\
        Average & 195465 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-1024} & 199185 & 199903 & 246245 & 247320 \\
        \hline
        \text{Kyber1024} & 222351 & 223260 & 258231 & 259067 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ Poly1305 & 205763 & 206499 & 51375 & 51562 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ GMAC & 208805 & 209681 & 54573 & 54780 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ CMAC & 213667 & 214483 & 59175 & 59408 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ KMAC & 216761 & 217468 & 62269 & 62516 \\
        \hline
    \end{tabular}
\end{table}
